Title of the report

### 2011-12-06 doc/report/report.tex
references in report

% check references! lncs is good maybe. lnai
\newpage
\chapter{Introduction}
\section{Introduction} 
>>This work is part of bigger project, spring and next spring.
>>Connection between grammar & parser: GF, every grammar defines a parser.
Ultimate goal: want computers to 'understand' language. Not good with flat strings.
We need to translate strings of natural language to a richer structure,
a parse tree. So far there is no
freely available ruled-based parsing giving a deep analyse for Swedish.
%There are parsers which put only put part-of-speach tags to
%the text, and others that gives a deeper analyse. 
There are statistical parser, and parser which only are deal with controlled language.
The GF parser are one of those. 
%What is parsing, flat strings to trees, example picture, what is deep parsing.
So far stable and good when working with controlled languages, but recent experiments
have showed that it is also possible to parse free language.
To do this for Swedish, we need to  .. the resource grammar needed to covering
the language specific constructions. Further a large lexicon is also needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
We have extended the Swedish resource grammar in GF, to give a better coverage and imported
a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & du & mig & inte?}
{Saw  & you&  me& not?}
{`Didn't you see me?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
%\enumsentence{
%\begin{tabular}{ll}
%{\shortex{7}
%{Jag & såg&  Anna.&  Hon & såg & inte&  mig.} 
%{ & & &  & & &}}
%\shortex{7}
%{Jag såg att Anna \textbf{inte} såg mig.}\\
%\shortex{7}
%{I  saw Anna. She saw not   me.}  & {I   saw that Anna not saw me}. \\
%`I saw Anna. $\;$ She did not see me.'& `I saw that Anna did not see me'.
%}
\enumsentence{Main sentences:\\
{\shortex{7}{Jag & såg & Anna. & Hon & såg & {\bf inte} & mig }
{ I & saw & Anna. & She & saw & {\bf not} & me}
{`I saw Anna. She didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & såg & att& Anna& \textbf{inte}& såg& mig.}
            {I &  saw & that& Anna& \textbf{not}& saw& me.}
            {`I saw that Anna didn't see me'}}

@@

\subsection{Grammatical Framework}
>> more here. First: grammar formalism with Martin-Löf.
>> snd: programming language, type checking and logic: prolog (check this)
>> dependent and implicit types : agda. (read gf book and krasses book)
>> pgf\ref{}, runtime system with parsing, lin, random generation osv.
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming languages
like Haskell\cite{haskell},
ML\cite{ml}, Prolog\cite{bratko} and
Agda\cite{agda}.
Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

@@

>>not very restricted, not few ambiguities.
>>the scope is controlled language, but possible ambiguities (discuss this in future
>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
>>syntactical (man with the telescope), 
>>lexical (youSg/Pl, växa).
A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt} and something else.
>> key, talk, hats, bom (see references in Ramonas thesis).
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
European project MOLTO\cite{molto} for online translation between up to 15 language.\\

@@

\vspace{5mm}
>> references here! to everything. Remove uncertain, add ontology, aarne paper about
>> donkey, find aarnes paper in ramona's thesis (french), hats formal methods,
>> bom.
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possibilities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics.
Further, experiments on using GF for free parsing has been conducted \cite[\textsection 4.1]{gfMech} 
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely available, manually annotated, large scale treebank,
Talbanken\cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
>> Much more here, important!! comparisons.
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happened to it. Logic and translation
to Swedish, The spoken lang. translator\cite{cle2}. Wiren \cite{wiren}, cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

For verbs, but also others there is an important distinction for valency.
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
Maybe talk about ställa.
Particles and prepositions, can't be seen on the trees.
tycka om /tycka om -> om vad tycker du inte. fast bra
about slashes\cite{gazdar}

>>>>

% check references! lncs is good maybe. lnai
\newpage
\chapter{Introduction}
\section{Introduction} 
>>This work is part of bigger project, spring and next spring.
>>Connection between grammar & parser: GF, every grammar defines a parser.
>> reusablilty, freely available.
>> the big amount of data, need language technology, machine translation
Ultimate goal: want computers to 'understand' language. Not good with flat strings.
We need to translate strings of natural language to a richer structure,
a parse tree. So far there is no
freely available ruled-based parsing giving a deep analyse for Swedish.
%There are parsers which put only put part-of-speach tags to
%the text, and others that gives a deeper analyse. 
There are statistical parser, and parser which only are deal with controlled language.
The GF parser are one of those. 
%What is parsing, flat strings to trees, example picture, what is deep parsing.
So far stable and good when working with controlled languages, but recent experiments
have showed that it is also possible to parse free language.
To do this for Swedish, we need to  .. the resource grammar needed to covering
the language specific constructions. Further a large lexicon is also needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
We have extended the Swedish resource grammar in GF, to give a better coverage and imported
a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & du & mig & inte?}
{Saw  & you&  me& not?}
{`Didn't you see me?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
>> and stricter
%\enumsentence{
%\begin{tabular}{ll}
%{\shortex{7}
%{Jag & såg&  Anna.&  Hon & såg & inte&  mig.} 
%{ & & &  & & &}}
%\shortex{7}
%{Jag såg att Anna \textbf{inte} såg mig.}\\
%\shortex{7}
%{I  saw Anna. She saw not   me.}  & {I   saw that Anna not saw me}. \\
%`I saw Anna. $\;$ She did not see me.'& `I saw that Anna did not see me'.
%}
\enumsentence{Main sentences:\\
{\shortex{7}{Jag & såg & Anna. & Hon & såg & {\bf inte} & mig }
{ I & saw & Anna. & She & saw & {\bf not} & me}
{`I saw Anna. She didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & såg & att& Anna& \textbf{inte}& såg& mig.}
            {I &  saw & that& Anna& \textbf{not}& saw& me.}
            {`I saw that Anna didn't see me'}}

@@

\subsection{Grammatical Framework}
>> more here. First: grammar formalism with Martin-Löf.
>> snd: functional and programming language, type checking and logic: prolog (check this)
>> dependent and implicit types : agda. (read gf book and krasses book)
>> pgf\ref{}, runtime system with parsing, lin, random generation osv.
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming languages
like Haskell\cite{haskell},
ML\cite{ml}, Prolog\cite{bratko} and
Agda\cite{agda}.
Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

@@

>>not very restricted, not few ambiguities.
>>the scope is controlled language, but possible ambiguities (discuss this in future
>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
>>syntactical (man with the telescope), 
>>lexical (youSg/Pl, växa).
A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY\cite{key}, in wich GF is used for translating specification
to natural language.
>> key, talk, hats, bom (see references in Ramonas thesis).
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
European project MOLTO\cite{molto} for online translation between up to 15 language.\\

@@

\vspace{5mm}
>> references here! to everything. Remove uncertain, add ontology, aarne paper about
>> donkey, find aarnes paper in ramona's thesis (french), hats formal methods,
>> bom.
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possibilities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics\cite{montague}.
Further, experiments on using GF for free parsing has been conducted
\cite[\textsection 4.1]{gfMech} 
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely available, manually annotated, large scale treebank,
>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
>> for large scale projects.
Talbanken\cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
>> Much more here, important!! comparisons.
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happened to it. Logic and translation
to Swedish, S-CLE (Gambäck,1997) full analyse but coverage is too limited (nivres papper), The
spoken lang. translator\cite{cle2}. Wiren \cite{wiren}. bit more information than
just pos: cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
SweCG (birn) annotation, parser Swedish FDG (voultanien,2001) which is not freely available.
finding subject, objects, suubject complements (SC) and
to link them to their proper regents (main verbs) : over 90\% both precision and recall.
uses a dependency grammar, but "The functional description of adverb phrases
and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
to be described in a future version." 
HPSG\cite{hpsg} and lingo matrix\cite{lingomatrix}, English, Japanese, German,
Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
(Ahrenberg) based on NorSource)
also Swedish: BiTSE\cite{stymne}, Stymne 2006, Swedish English translation,
focusing on verb phrames, built on HPSG.
S\o rgaard and Haugereid, Scandinavian Grammar Matrix\cite{scandmatrix}
Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English: ongoing to get widecoverage grammar
under GPL. Translation to Korean.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

For verbs, but also others there is an important distinction for valency.
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
Maybe talk about ställa.
Particles and prepositions, can't be seen on the trees.
tycka om /tycka om -> om vad tycker du inte. fast bra
'jag tar med mig den till Bo' : ta\_med\_V3 = dirV3 (reflV (partV (take\_V "med")) (mkPrep "till") ;
but is really 'till' part of verb?
VP is divided corresponding to satsschema (diderichsen) (renewed by SAG),
otherwise use slash.
"Speed up", instead of movement used slashes\cite{gazdar}.
Slashes good for question (vilken katt ser du idag)
relatives (katten som du tittar på) and topicalisation
The grammar has VP,Cl, SSlash, all missing NP.
